---
title: "Assignment 5: Substance Use Treatment Mapping"
subtitle: "Lars Richel"
author: "Amber Camp"
date: "`r Sys.Date()`"
format: html
editor: visual
toc: true
---

## Overview

Due: **Friday 10/2, 11:59pm**

In this assignment, you will apply analytic and mapping skills to the 2024 TEDS-D dataset from SAMHSA.

Refer to the following files for reference, though some of what is asked of you here will require logic beyond what we discussed in class:

- week06-treatment-mapping-part1.qmd

- week06-treatment-mapping-part2.qmd

**Before you start:** keep the [codebook](https://www.samhsa.gov/data/system/files/media-puf-file/TEDS-D-2024-DS0001-info-codebook_v1_0.pdf) open in another tab. Every question below depends on knowing what the values in a column actually mean. Don't assume — look it up.

**Time estimate:** about three-four hours total. Question 1 should take you ten minutes. If you are stuck on one thing for more than twenty minutes, ask your classmates or get in touch with me. Being stuck is normal. Staying stuck quietly by yourself is not :)

## Load Packages and Data

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(sf) # "simple features" for spatial data
library(urbnmapr) # access to US state/country shapefiles for mapping
library(ggiraph) # makes ggplot charts interactive
options(scipen = 99) # turns off scientific notation
```

This is the cleaned dataset you saved at the end of Part 1.

```{r}
#| label: read-data

teds_d <- readRDS("data/teds_d_clean.rds")
```

## Question 1: Warm-up

Before mapping anything, get oriented in the data.

a)  Using `tabyl()`, show the breakdown of `reason` (reason for discharge). Which reason is most common, and what percentage of all discharges does it account for?

b)  Make one bar chart showing the number of discharges by `services_d` (type of treatment service). Give it an informative title and axis labels.

c)  In a sentence or two: what does the `reason` value "Transferred to another treatment program or facility" mean for our completion percentages? If a state transfers a lot of patients, would its completion rate look higher or lower than a state that transfers very few?

> **Full credit:** a working `tabyl()`, a readable labeled bar chart, and a sentence or two showing you thought about part (c). There is no trick in this question!

```{r}
#| label: q1
teds_d %>%
  tabyl(reason)

ggplot(teds_d, aes(x = services_d)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Number of Discharges by Treatment Service Type",
    x = "Service Type",
    y = "# of Discharges"
  ) + 
  theme_minimal() +
  coord_flip()
```

**Answer to (a): Treatment Completed 48%**

**Answer to (b):** (produce a bar chart in the code chunk above)

**Answer to (c): It means that the completion rate will be lower because they did not complete the treatment at that facility. So, a state that transfers a lot of patients would appear to have a lower completion rate than one that transfers very few.**

## Question 2: Mapping no use at discharge

Make **one [*interactive*]{.underline} map** with `ggiraph` showing, for each state, the percentage of [completed treatment]{.underline} episodes that ended with [no use at discharge]{.underline}.

In other words: among the episodes where `reason` is "Treatment completed," what share had `freq1_d` equal to "no use"?

Work through this in bite-sized steps, like we did in Part 2. You will need to filter, group, summarize, build an index, join, and plot. Write one step at a time and look at the result before moving on.

### The denominator decision

About a third of the `freq1_d` values in this dataset are "unknown." That means you have to decide what your denominator is, and there is more than one defensible answer:

- **All** completed episodes, counting "unknown" in the denominator, or

- Only completed episodes where frequency of use was **actually recorded**.

Pick one. **Below your map, write one sentence saying which denominator you used and why.** Either choice can earn full credit. Missing description earns no points for this portion.

### Your map must:

- be interactive (rendered with `girafe()`)

- show a tooltip with the state name and the percentage, rounded to two decimal places

- have a title and a legend label that say what is being shown

- handle states with no data honestly (remember Friday in class)

> **Full credit:** one interactive map meeting the four requirements above, plus your denominator sentence. It does not have to be beautiful, only correct and readable.

```{r}
#| label: q2
no_use_percent <- teds_d |>
  filter(reason == "Treatment completed") |>
  filter(freq1_d != "unknown") |>
  group_by(stfips) |>
  summarize(
    total_completed = n(),
    no_use_count= sum(freq1_d == "no use", na.rm = TRUE),
    pct_no_use = round(100 * no_use_count / total_completed, 1)
  )

states_map <- get_urbn_map(map = "states", sf = TRUE) |>
  mutate(stfips = as.numeric(state_fips))

map_data <- states_map |>
  left_join(no_use_percent, by = "stfips")

gg_map <- ggplot(map_data) + 
  geom_sf_interactive(
    aes(
      fill = pct_no_use,
      tooltip = paste0(state_name, ": ", pct_no_use, "% no use at discharge"),
      data_id = state_fips
    ),
    color = "white",
    size = .2
  )+
  scale_fill_viridis_c(
    name = "% No Use",
    option = "viridis",
    na.value = "grey"
  ) +
  labs(
    title = "Percentage of Completed Treatment Episodes Ending in No Use",
    subtitle = "By State"
  ) +
  theme_void()

girafe(ggobj = gg_map)
```

**Denominator I used, and why: I used competed treatments where frequency was not unknown because including those cases would lower the percentage for states with higher rates of missing data.**

## Question 3: Service type and length of stay

Treatment outcomes can vary depending on the type of service a patient receives and how long they stay in treatment.

Investigate how **completion rates** differ across **service type** (`services_d`) and **length of stay** (`los`).

- Create **two** visualizations: at least one focused on service type, and at least one focused on length of stay. These do not need to be maps, and they do not need to be interactive. Pick whatever chart type makes the pattern easiest to see.

- Then write a short summary (3-ish sentences, full grammatical sentences) explaining the main patterns you observe.

**One warning:** look up `los` in the codebook *before* you calculate anything with it. Code that runs without an error can still give you a meaningless answer.

> **Full credit:** two labeled visualizations that actually show the comparison you say they show, plus a summary that describes a real pattern in your own output. You are not graded on finding the "right" pattern. You are graded on whether your summary matches your charts.

```{r}
#| label: q3
service_completion <- teds_d |>
  filter(!is.na(services_d), !is.na(reason)) |>
  group_by(services_d) |>
  summarize(
    total = n(),
    completed_count = sum(reason == "Treatment completed", na.rm = TRUE),
    completion_rate = completed_count / total
  )

ggplot(service_completion, aes(x = reorder(services_d, completion_rate), y =completion_rate)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip()+
  labs(
    title = "Treatment Completion Rate by Service Type",
    x = "Service Type",
    y = "Completion Rate"
  ) + 
  theme_minimal()

los_completion <- teds_d |>
  filter(!is.na(los), !is.na(reason)) |>
  group_by(los) |>
  summarize(
    total = n(),
    completed_count = sum(reason == "Treatment completed", na.rm = TRUE),
    completion_rate = completed_count / total
  )

ggplot(los_completion, aes(x = factor(los), y = completion_rate)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(
    title = "Treatment Completion Rate by LOS",
    x = "LOS",
    y = "Completion Rate"
  ) +
  theme_minimal()
```

**Summary: Completion Rates vary across service types and length of stay. In service types, the highest average completion rates come from detoxification programs, while residential programs fall in the middle, and outpatient programs coming in last. However, in length of stay there is no real correlation. You cannot be confident in saying that the longer you stay the more likely you are to complete treatment, and you cannot say that the shorter you stay the more likely you are to complete treatment.**

## Extra Credit (optional, up to 2 pts)

#### Again, this is meant to *stretch* you, not *stress* you!

Make one visualization that shows service type and length of stay **together**–that is, how completion rate varies across both at once. Faceting is one way, but there are others.

Then, in a sentence, say what this combined view shows that your two separate charts in Question 3 could not.

```{r}
#| label: extra-credit
comp <- teds_d |>
  filter(!is.na(services_d), !is.na(los), !is.na(reason)) |>
  group_by(services_d, los) |>
  summarize(
    total = n(),
    completed_count = sum(reason == "Treatment completed", na.rm = TRUE),
    completion_rate = completed_count / total, 
    .groups = "drop"
  )
ggplot(comp, aes(x= factor(los), y =completion_rate)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~services_d) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Completion Rate by LOS Across Service Types",
    x = "LOS",
    y = "Completion Rate"
  )+ 
  theme_minimal()
```

**What the combined view adds: This shows there is a correlation between length of stay and completion rate, it just depends on what service type you are looking at. Some service types show no correlation, but some also show that the longer you stay the more likely you are to complete treatment.**

------------------------------------------------------------------------

## Push to GitHub AND submit link on Canvas!

Assignment is due **Friday 10/2, 11:59pm**.
